---
title: "Lab #7: Components of AI (30 pts)"
author: "Steph Krieger"
format:
  html:
    code-fold: false
execute: 
  echo: true
knitr:
  opts_chunk: 
    message: false
editor: source
editor_options: 
  chunk_output_type: inline
---

## Introduction

As you've seen in this week's lectures, much forethought has to go into the creation of a model for artificial intelligence, whether it is from the business, engineering, or ethical side of things. We will focus on the engineering and ethical portions by using a poorly-made example of a natural language processor using Markov chains.

Before we dive into the mechanics of the model, discuss with others around you and write answers regarding the following questions:

::: {.callout-important}
###### Q1 (3 pts)

Think about the following sentence: "I am having a great day!". What rules or considerations do I need to follow linguistically in order to convey that sentiment? Is there an order to how those words are being said?

One rule you need to follow when reading in order to convey the sentiment is to read each word in order. The question mark also conveys how it is an excited statement. 

:::

::: {.callout-important}
###### Q2 (3 pts)

Think about the following sentence: "I am having a great day!". Are there replacements I can make for each word that would still result in a comprehensible sentence?

One way you could make replacements is to change the words to "Today is a great day!" or you could also change the sentence to past tense "I had a great day!" 

:::

::: {.callout-important}
###### Q3 (3 pts)

Think about the following sentence: "I am having a great day!". How often am I using each of these words in a sentence? Does it depend on the situation?

Each word is used once. But sometimes a word can be used more than once depending on the sentence. 

:::

## Behind the AI

Despite over-analyzing the six-worded sentence, it is important to understand the mechanics of this sentence generation when it comes to generating responses. A large-language model (LLM) is a computational model that uses a bunch of data for natural language processing. When we build a model, we want to consider the appropriate methods that can be used for training and results, as well the inputs and outputs associated with our method.

Pivoting back to our discussion on how sentences are created, one model we might create builds upon the idea of predicting what the next word might be given the word before, based on the part of speech and subject. Here, we venture into using [Markov chains](https://en.wikipedia.org/wiki/Markov_chain). 

We will need to assign probabilities (and/or weights) to how the next word of a sentence comes about. Additionally, we'll need a dictionary of words to start with.

There are many sophisticated models that can be used for natural language processing, including but not limited to:

- Deep learning models
- Neural networks
- Large-language models

We will not delve into these specific models; rather, we will focus on a more simplistic model to garner a better understanding of how a probabilistic model can fuel generation of a sentence.

I have created two models to aid in our creation of sentences:

- A model to generate probabilities of certain words being used, given their part of speech (i.e. noun, adjective)
- A model to generate probabilities of certain parts of speech being used next given the part of speech from the previous word

These two models allow for the determination of a word given what part of speech we're using and what the next word might be given what word we've used. As you'll see in a second, this approach is rather interesting.

To incorporate these two models and generate sentences, we will use two different datasets:

- `wordData.csv`: This contains 114 different words with the part of speech they fall under, their usage probability, and the chance of which part of speech comes next.
  - Words come from the list of most commonly used words in English (Source: [Wikipedia](https://en.wikipedia.org/wiki/Most_common_words_in_English))
- `tranMatData.csv`: This is a transition matrix that allows us to go from one word to the next.

A transition matrix can be thought of as a grid of outcomes/chances depending on the action. A small example is to determine what mood you'll be in the next day given how you feel currently. If you're in a bad mood today, perhaps there's a 60\% chance you'll be in a good mood tomorrow. On the flip side, if you're in a good mood today, perhaps there will be a 45\% chance you'll be in a good mood tomorrow. When placed in a transition matrix, we get the following:

|        | Happy | Sad  |
|:-------|:------|:-----|
| Happy  | 0.45  | 0.55 |
| Sad    | 0.60  | 0.40 |

In our example, we are using the transition matrix to go from one word to the next. Load in the [wordData.csv](wordData.csv) and [tranMatData.csv](tranMatData.csv) files and then take a look at the `head()` of the two data files.

```{r}
wordData <- read.csv("wordData.csv")
tranMatData <- read.csv("tranMatData.csv")

head(wordData)
head(tranMatData)
```

Answer the following questions below:

::: {.callout-important}
###### Q4 (3 pts)

Look at `wordData`. The `prob` column provides the probability of that word being used given what its part of speech is. Determine if these probabilities are legitimate for each of these words. What other considerations do I need to make when determining the probability?

I think these probabilites are legitimate for these words. However, one consideration that should be taken into account when determining probability is the context of each sentence. For example, some words may have relatively small probabilities of being used in every day conversation, like government (0.72%), but if a person is specifically in a conversation about the government, the probability of that word being used drastically increases. 

:::

::: {.callout-important}
###### Q5 (3 pts)

Look at `tranMatData`. Each row provides the probability that the corresponding word will be followed by the word specified by the column. Determine if these probabilities are legitimate for each of these words. What other considerations do I need to make when determining the probability?

I think these probabilities are legitimate because to me it makes sense that words like small, large, or the are more popular than words like person, or way. however, one consideration I think should be made when determining probability is again the context of the conversation. I feel like the probability of certain words coming up greatly depends on the topic being discussed.

:::

::: {.callout-important}
###### Q6 (3 pts)

Based on both of the datasets, what are some issues that may arise when using this data?

Some issues that may arise when using this data is not enough diversity in words for specific conversations. If only the popular words are coming up and not specific words about different topics than it can lead to sentences that don't fully make sense in the context of different conversations. Additionally it looks like the words with the highest probabilities are all filler words, which will lead to sentences that just have no real meaning.

:::

## Running the Model

Now onto the fun part -- we are going to create our own sentences! To do so, run the following chunk below:

```{r}
# Code to provide the first word of the sentence
sentence = c("the")
wordInd = which(wordData$word == "the")

# We can specify the number of words we want in our sentence here
numWords = 6

# Run this code to create the sentence
for (i in 1:(numWords - 1)){
  wordInd = sample(1:nrow(wordData), size = 1, prob = tranMatData[wordInd, ])
  sentence = c(sentence, wordData$word[wordInd])
}

# Print the sentence here
paste(sentence, collapse = " ")
```

There you have it -- we've generated sentences! But as you see, the results are interesting to say the least. You'll notice that every time you run it, you will generate a new sentence. Some may make sense, while others may not. 

As you've probably discussed in previous questions, there are many, many considerations to generating these sentences. The following questions will bring the discussion back full-circle:


::: {.callout-important}
###### Q7 (3 pts)

Run the code above 5 different times using a `numWords` of 6. Copy and paste the results below.

1. "the person to person get man"
2. "the time that life to other"
3. "the year have woman say world"
4. "the year and point say company"
5. "the year be day go way"
6. "the person know time be way"


:::

::: {.callout-important}
###### Q8 (3 pts)

What are some observations you notice from the sentences? Are there any similarities? Differences? Confusions? Do the sentences make sense?

The sentences all make no sense. They are just full of filler words with no real context. One observation was 'year' was a word that came up many times. 

:::

::: {.callout-important}
###### Q9 (3 pts)

What are some factors we may want to factor into our program aside from:

- the probability of a word being used, and
- what the next word will be given the part of speech of the previous word?

I think just focusing on adding more words of meaning to the program. Maybe making a program that instead of seeing what the next word will be given the part of speech of the previous word it instead looks at what the next word will be given the meaning of the previous word. This way the sentences will make more sense and express full ideas rather than just being random collections of filler words. 

:::

::: {.callout-important}
###### Q10 (3 pts)

Suppose we attempted to use this model in lieu of more sophisticated programs like ChatGPT, Gemini, and Claude. What would your perception be? How would this lead to issues with resolving prompts you fed to the programs?


I think if this model was how ChaptGPT, Gemini, or Claude was it would be a lot less popular and less used, especially for help on academic related things. I would bet instead of feeding academic prompts to these forms of AI people would just do them themselves because I think using this model would just make it more confusing to complete a task. Maybe making AI models like this is a good strategy to get less people to rely on them becuase this is just far less helpful.
:::

## Wrapup

Thankfully, these sophisticated programs are able to answer questions and provide responses that answer our questions, do so in a natural way, and give additional information aside from the initial ask. This is what has made these programs so intriguing and highly used over the last few years.

However, as you've seen with this example, developing these models takes a lot of foresight and power. The factors you've discussed in this activity only scratch the surface of all potential variables and factors to consider when developing a model. Additionally, we only looked at prompt generation and not into other avenues, such as comprehending prompts. 

If you've already exhausted a lot of brain power thinking about these factors, imagine the power it takes to factor these considerations into a series of models and executing the results. The data itself will also compound these actions. Also, as you've seen here, artificial intelligence relies on the success of the models, data, and processes behind the scenes. Nothing is perfect, and at times, you may find that these models will produce odd, incomplete, or unrelated answers.

## Prompt

In Slack (under the `#discussions` channel), provide an example of a time you used an large language model like ChatGPT to resolve a prompt and it generated a rather unusual response. Consider why the response was generated as such.

  