---
title: "Lab #9: Mapping with Leaflet"
author: "Steph Krieger"
format:
  html:
    code-fold: false
execute: 
  echo: true
knitr:
  opts_chunk: 
    message: false
editor: source
editor_options: 
  chunk_output_type: inline
---


## Setting Up Your Environment

1. Install the following packages in your Environment:

-   `install.packages("leaflet")`
-   `install.packages("sf")`

2. Run the code below to load today's data frames into your environment. 

```{r}
#| document: show
#| warning: false
library(tidyverse)
library(leaflet)
library(RColorBrewer)
library(sf)

tri_la_2023 <- read_csv("https://raw.githubusercontent.com/SDS-192-Intro/public-website-fall-25/refs/heads/main/data/2023_la.csv", name_repair = make.names) |>
  filter(X50..UNIT.OF.MEASURE == "Pounds") |>
  arrange(desc(X107..TOTAL.RELEASES)) |>
  mutate(chemical_total_text = paste(X37..CHEMICAL, X107..TOTAL.RELEASES, sep = ": ")) |>
  group_by(X2..TRIFD) |>
  summarize(X4..FACILITY.NAME = first(X4..FACILITY.NAME),
            X12..LATITUDE = first(X12..LATITUDE),
            X13..LONGITUDE = first(X13..LONGITUDE),
            X23..INDUSTRY.SECTOR = first(X23..INDUSTRY.SECTOR),
            X107..TOTAL.RELEASES = sum(X107..TOTAL.RELEASES),
            X..CHEMICAL.TEXT = paste(chemical_total_text, collapse = " <br> "))
```

## Questions

::: {.callout-important} 
###### Question 1 (3 pts)

Adjust the code below to center the map on Louisiana. You'll need to look up the coordinates for LA, keeping in mind that South and West coordinates will be negative. Adjust the `zoom` to keep the whole state in view (setting the zoom level between 1 and 16). When you are happy with the View, switch out the provider tiles to a base map that won't distract from the data points we will layer on top of this map. Keep in mind our discussions regarding *Visualization Aesthetics* here. 

```{r}
# Uncomment and adjust code below.
la_map <- leaflet(width = "100%") %>%
  setView(lat = 29.98, lng = -92.09, zoom = 5.5) %>%
  addProviderTiles("CartoDB.Positron")

# Uncomment below to view the map!
 la_map
```

:::

::: {.callout-important} 
###### Question 2 (3 pts)

In the code block below, create a new `geometry` column with the given coordinates, using the function `st_as_sf()`. The column for longitude will go in the first coordinate position, and latitude will go in the second. Be sure to set the data's current CRS (4269) in that function. 
Then transform the CRS to 4326 using `st_transform()`. 

```{r}
# Uncomment and fill in the blanks below. The column for longitude will go in the first coordinate position, and latitude will go in the second. 

tri_la_2023 <- tri_la_2023 %>%
  st_as_sf(coords = c("X13..LONGITUDE", "X12..LATITUDE"), crs = 4269) %>%
  st_transform(4326)
```

:::

::: {.callout-important} 
###### Question 3 (3 pts)

In the code block below, pipe `addCircleMarkers()` onto the basemap and list `data = tri_la_2023` as an argument in that function. It should look like my map below. 


```{r}
# Uncomment and add the function

la_map %>% leaflet(width = "100%") %>%
  setView(lat = 29.98, lng = -92.09, zoom = 5.5) %>%
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = tri_la_2023)
```

:::

::: {.callout-important} 

###### Question 4 (3 pts)

Copy the map you just created into the code chunk below, and pipe the `addCircleMarkers()` function onto your code. Check out the help pages for the `addCircleMarkers()` function, and add some arguments to help with the map's legibility. At the very least, you should adjust the `radius`, `weight`, `color`, `fillColor`, and `fillOpacity`, and understand how each of these arguments will change the style of the map. For now you can set the `color` to "black" and the `fillColor` to "white". See if you can get your map to match mine below. 

```{r}
# Copy previous map here!

la_map %>% leaflet(width = "100%") %>%
  setView(lat = 29.98, lng = -92.09, zoom = 5.5) %>%
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = tri_la_2023,
                   radius = 2,
                   color = "black",
                   weight = 0.6,
                   fillColor = "white",
                   fillOpacity = 0.6)
```

:::


::: {.callout-important} 
###### Question 5 (3 pts)

Copy and paste the last map that you created below. We are going to size each circle by the total chemical releases at that facility. Total releases is stored in the column `X107..TOTAL.RELEASES` in our dataset, so we could size the circles by setting the radius to `~X107..TOTAL.RELEASES`. However, this would lead to some massive circles on our map as certain facilities have released hundreds of thousands of pounds in emissions, and the value we supply to radius will determine the pixels of the circle on our map. To deal with this, we are going to divide the total releases by 1000000. Set the radius to `~X107..TOTAL.RELEASES/1000000` below.

```{r}
# Copy map colored by bins here!

la_map %>% leaflet(width = "100%") %>%
  setView(lat = 29.98, lng = -92.09, zoom = 5.5) %>%
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = tri_la_2023,
                   radius = ~X107..TOTAL.RELEASES/1000000,
                   color = "black",
                   weight = 0.6,
                   fillColor = "white",
                   fillOpacity = 0.6) 
```

:::

::: {.callout-important} 
###### Question 6 (3 pts)

Create a categorical palettes below using `colorFactor()`, setting the `palette` to "Set2", and the domain to `tri_la_2023$X23..INDUSTRY.SECTOR`. 

```{r}
#Uncomment below to create the palette

pal_fact <- colorFactor(palette = "Set2", 
                        domain = tri_la_2023$X23..INDUSTRY.SECTOR,)


```

:::


::: {.callout-important} 
###### Question 7 (3 pts)

Copy and paste the last map you created into the code chunk below. We are going to adjust the `fillColor` by setting it to the variable from the dataset that we wish to color (`X23.INDUSTRY.SECTOR`), colored by one of the palettes that we created. We can do this by setting the fill color argument equal to `~pal_fact(X23.INDUSTRY.SECTOR)`. 

We also need to add a legend to explain the colors. Add a pipe to the end of the `addCircleMarkers()` and then add the function `addLegend()`. Consult the help pages for the `addLegend()` function to determine how to add a legend for the meaning of the colors represented on the map. At the very least, you will need an argument for `data`, `pal`, and `values`.

Your map should look like mine below. 

```{r}
# Copy the map here!

la_map %>% leaflet(width = "100%") %>%
  setView(lat = 29.98, lng = -92.09, zoom = 5.5) %>%
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = tri_la_2023,
                   radius = ~X107..TOTAL.RELEASES/1000000,
                   color = "black",
                   weight = 0.6,
                   fillColor = ~pal_fact(X23..INDUSTRY.SECTOR),
                   fillOpacity = 0.6) |>
addLegend(data = tri_la_2023,
          title = "TRI Releases in Pounds",
            pal = pal_fact, 
            values = ~X23..INDUSTRY.SECTOR)
  
```

:::

::: {.callout-important} 
###### Question 8 (3 pts)

Create three palettes below (one using each function `colorNumeric()`, `colorBin()`, and `colorQuantile()`), setting the `palette` to "YlOrRd", and the domain to `tri_la_2023$X107..TOTAL.RELEASES`. For `colorBin()`, set the `bins` to 6, and for `colorQuantile()`, set `n` to 4.

```{r}
pal_num <- colorNumeric(palette="YlOrRd", 
                        domain = tri_la_2023$X107..TOTAL.RELEASES)
#Uncomment below to create the other two palettes.

pal_bin <- colorBin(palette = "YlOrRd",
                    domain = tri_la_2023$X107..TOTAL.RELEASES,
                    bins = 6)

pal_quant <- colorQuantile(palette = "YlOrRd",
                    domain = tri_la_2023$X107..TOTAL.RELEASES,
                    n = 4)

```

:::

::: {.callout-important} 
###### Question 9 (3 pts)

Copy and paste the last map you created into the code chunk below, three times. For each, we are going to adjust the `fillColor` by setting it to the variable from the dataset that we wish to color (`X107..TOTAL.RELEASES`), colored by one of the palettes that we created. We can do this by setting the fill color argument equal to:

-   `~pal_num(X107..TOTAL.RELEASES)`: for coloring the points according the `pal_num()` palette we created on the `X107..TOTAL.RELEASES` variable
-   `~pal_bin(X107..TOTAL.RELEASES)`: for coloring the points according the `pal_bin()` palette we created on the `X107..TOTAL.RELEASES` variable
-   `~pal_quant(X107..TOTAL.RELEASES)`: for coloring the points according the `pal_quant` palette we created on the `X107..TOTAL.RELEASES()` variable

Adjust the legends for each of the maps.

Your maps should look like mine below. 

```{r}
# Copy map first time here!

la_map %>% leaflet(width = "100%") %>%
  setView(lat = 29.98, lng = -92.09, zoom = 5.5) %>%
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = tri_la_2023,
                   radius = ~X107..TOTAL.RELEASES/1000000,
                   color = "black",
                   weight = 0.6,
                   fillColor = ~pal_num(X107..TOTAL.RELEASES),
                   fillOpacity = 0.6) |>
addLegend(data = tri_la_2023,
          title = "TRI Releases in Pounds",
            pal = pal_num, 
            values = ~X107..TOTAL.RELEASES)

# Copy map second time here!

la_map %>% leaflet(width = "100%") %>%
  setView(lat = 29.98, lng = -92.09, zoom = 5.5) %>%
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = tri_la_2023,
                   radius = ~X107..TOTAL.RELEASES/1000000,
                   color = "black",
                   weight = 0.6,
                   fillColor = ~pal_bin(X107..TOTAL.RELEASES),
                   fillOpacity = 0.6) |>
addLegend(data = tri_la_2023,
          title = "TRI Releases in Pounds",
            pal = pal_bin, 
            values = ~X107..TOTAL.RELEASES)

# Copy map third time here!

la_map %>% leaflet(width = "100%") %>%
  setView(lat = 29.98, lng = -92.09, zoom = 5.5) %>%
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = tri_la_2023,
                   radius = ~X107..TOTAL.RELEASES/1000000,
                   color = "black",
                   weight = 0.6,
                   fillColor = ~pal_quant(X107..TOTAL.RELEASES),
                   fillOpacity = 0.6) |>
addLegend(data = tri_la_2023,
          title = "TRI Releases in Pounds",
            pal = pal_quant, 
            values = ~X107..TOTAL.RELEASES)
```

:::

::: {.callout-important} 
###### Question 10 (3 pts)

Below explain the following: Why do the colors appear differently on each map? Which map best represents the distribution of values in `X107..TOTAL.RELEASES`?

The colors appear differently on each map because on each map the color scale is assigned to different levels/numbers. In my opinion, the map that best represents the distribution of values in 'X107..TOTAL.RELEASES' is the pal_bin one because it shows a scale with the most range allowing for the most amount of colors (and pounds of TMI release) to be displayed. I think this is best because it shows specifically how many pounds were released in each area instead of categorizing the amount of pounds into buckets.

:::

::: {.callout-important} 
###### Question 11 (3 pts)

Now, we are going to add popups on the map. Copy and paste the code from the previous step into the code chunk below. Then do the following:

1. Add the `popup` argument to the `addCircleMarkers()` function. Using the `paste()` function, create a popup that includes the facility's name (in bold), the facility's industry sector (in italics), and the chemicals released by the facility. (Note that the last the data point is stored in `X..CHEMICAL.TEXT`). Note that you will need to place a `~` in front of the `paste()` function to communicate that there are variables in your argument. 

2. Add the `popupOptions` argument to the `addCircleMarkers()` function. Set the `popupOptions` to `popupOptions(maxHeight = 200, closeOnClick = TRUE))`. This will ensure that really tall popups will scroll vertically rather than taking up an entire page. 

Your popups should look like mine in the map below.


```{r}
# Copy previous map here!

la_map %>% leaflet(width = "100%") %>%
  setView(lat = 29.98, lng = -92.09, zoom = 5.5) %>%
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = tri_la_2023,
                   radius = ~X107..TOTAL.RELEASES/1000000,
                   color = "black",
                   weight = 0.6,
                   fillColor = ~pal_quant(X107..TOTAL.RELEASES),
                   fillOpacity = 0.6,
                   popup = ~paste("<b>",X4..FACILITY.NAME, "</b><br><i>", X23..INDUSTRY.SECTOR, "</i><br><br>", X..CHEMICAL.TEXT)) |>
addLegend(data = tri_la_2023,
          title = "TRI Releases in Pounds",
            pal = pal_quant, 
            values = ~X107..TOTAL.RELEASES)
```

:::

::: callout-caution
## Ethical Considerations

Notably, while the Emergency Planning and Community Right-to-Know Act mandates *reporting* of emissions, it does not mandate *monitoring* of emissions. While other environmental regulations do set certain monitoring standards for specific TRI chemicals and pollution activities, for all other chemicals and activities, facilities are required to report based on a "reasonable estimate" of releases and other waste management quantities. In the early 1990s the National Wildlife Federation uncovered that certain facilities showing tremendous reductions in emissions from one year to the next had not actually cut down on pollution, but instead developed different ways of estimating their emissions. The public doesn't see how emissions are calculated; they only see the final reported emissions. Given this history of certain facilities reporting "phantom reductions", what kinds of legal, practice-based, or social interventions might ensure that this data remains reliable and accountable to the communities that use it to assess risks?  What might make these interventions challenging to implement; what would need to change in order for these interventions to be realized? Share your ideas on our `discussions` Slack channel.

:::

